Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-01-16 HickernellMCQMC2024_QMCTutorial.tex
Updates from Overleaf

%When optimizing a sample point distribution, one requires a figure-of-merit to serve as the objective function for the optimization problem.

\vspace{2mm}
\noindent
\textbf{Combinatorial Optimization.}
For several decades, combinatorial optimization techniques have been employed to obtain good generating vectors for rank-1 lattice points. Recall that for given \(n\) and \(d\), a lattice rule is fully determined by the generating vector \(\boldsymbol{h} \in \mathbb{Z}_n^d \coloneqq \{0, 1, \ldots, n-1\}^d\). While good generating vectors are known for \(d \leq 2\) \cite{DicEtal22a,SloJoe94}, the same does not hold for \(d \geq 3\), often resorting to finding good $\boldsymbol{h}$ via computer search to minimize a chosen figure-of-merit. An exhaustive search is infeasible due to the exponential growth of the search space with \(d\). The component-by-component (CBC) construction, introduced by Korobov \cite{kor63} and later revisited by Sloan and collaborators \cite{Slo02}, reduces the search space to size \(dn\) by constructing \(\boldsymbol{h}\) one component at a time. 
%Further simplification is possible by restricting the search to Korobov lattices of the form \(\boldsymbol{h} = (1, h, h^2, \ldots, h^{d-1}) \in \mathbb{Z}^d\). 
For certain choices of function spaces, there exist fast component-by-component construction implementations using fast Fourier transforms (FFTs) as the primary tool. We refer to \cite{DicKuo04a,JoeKuo03,KuoJoe02b} and references therein for works on the CBC construction for lattice rules and \cite{NuyCoo06a,NuyCoo06b} for their fast implementations.

\vspace{2mm}
\noindent
\textbf{Global Optimization.}

\vspace{2mm}
\noindent
\textbf{Other Notable Methods.}

@@

\subsection{Digital Scrambles} \label{sec:scrambles}
For digital nets one can randomize even further by a scrambling that preserves the $t$-value.  Owen \cite{Owe95} proposed a scrambling of nets called nested uniform scrambling.  A simpler version called linear matrix scrambling was proposed by Matou\v{s}ek \cite{Mat98} and is described here.  Starting from the formulation of digital sequences involving generator matrices in \ref{eq:digitalB}, we multiply each generator matrix on the left by a lower triangular matrix with ones along the diagonals and elements below the diagonal that are randomly chosen to be $0$ or $1$ with equal probability:
\begin{subequations} \label{eq:scr_lattice}
\begin{gather} \label{eq:digitalscr}
	\begin{pmatrix} x^{\text{scr}}_{ij1} \\ x^{\text{scr}}_{ij2} \\ \vdots \end{pmatrix}
	= \mathsf{L_j} \mathsf{C}_j \begin{pmatrix} i_0 \\ i_1 \\ \vdots \end{pmatrix} + \begin{pmatrix} \Delta_{1j} \\ \Delta_{2j} \\ \vdots \end{pmatrix} \quad \pmod{2}, \qquad j = 1,\ldots, d, \ i = 0, 1, \ldots, \\
	\mathsf{L_j} :=
	\begin{pmatrix}
		1 & 0 & 0 & 0 & \cdots \\
		l_{21} & 1 & 0 & 0 & \cdots \\
		l_{31} & l_{32} & 1 & 0 & \cdots \\
		\vdots & \vdots & \vdots & \ddots & \ddots
	\end{pmatrix}, \qquad l_{\ell j} \overset {\text{IID}}{\sim} \calU\{0,1\}
\end{gather}
\end{subequations}
A random digital shift, $\bsDelta$ is also added, where $\Delta_{\ell j}$ denotes the $\ell^{\text{th}}$ digit of the $j^{\text{th}}$ of the shift.

@@

@@

\section{Discrepancy} \label{sec:discrepancy}
So far, we have relied on eye tests and the Keister example to show how LD sequences are better than IID.  This section introduces the theory that shows way minimizing discrepancy leads to better approximations to $\mu$.

@@

Having a reproducing kernel Hilbert space with a known kernel $K$ allows us to derive a rigorous error bound for $\mu - \hat{\mu}_n$.  First note that for any bounded, linear functional, $L$ on the Hilbert space $\calH_K$, there is some $\eta_L \in \calH_K$ such that
\begin{equation*}
L(f) = \langle \eta_L , f \rangle_{\calH_K} \qquad \forall f \in \calH_K.
\end{equation*}
This is guaranteed by the Riesz Representation Theorm, and $\eta_L$ is called the representer of $L$.  Using the reproducing property of $K$, one may derove am explict formula for $\eta_L(\bsx)$, namely,
\begin{equation*}
\eta_L(\bsx) = \langle K(\cdot, \bsx), \eta_L \rangle_{\calH_K} = L\bigl( K(\cdot,\bsx) \bigr).
\end{equation*}
From this expression for $\eta_L(\bsx)$ one may then calculate the squared norm of the linear functional $L$ as the squared norm of its representer:
\begin{equation} \label{eq:sqnormL}
	\lVert \eta_L \rVert_{\calH_K}^2 = \langle\eta_L, \eta_L \rangle_{\calH_K} = L(\eta_L) = L^{\cdot\cdot} \Bigl(L^{\cdot}\bigl( K(\cdot,\cdot \cdot) \bigr) \Bigr).
\end{equation}

If $\mu(\cdot)$ is a bounded linear functional on $\calH_K$ is, we may use the argument above to derive our error bound.  First we write the dependence of $\mu - \hat{\mu}_n$ on  $f$ explicitly as $\mu(f) - \hat{\mu}_n(f)$ and note that the error functional, $\mu(\cdot) - \hat{\mu}_n(\cdot)$, is  linear.  The Riesz Representation Theorem and the Cauchy-Schawarz inequality imply a tight error bound:
\begin{align}
	\nonumber
	\lvert\mu(f) - \hat{\mu}_n(f) \rvert
	& = \lvert \langle \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)}, f \rangle_{\calH_K} \rvert
	 \le \lVert  \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)} \rVert_{\calH_K} \, \lVert f \rVert_{\calH_K}\\
	\lvert\mu(f) - \hat{\mu}_n(f) \rvert
	& \le \underbrace{\lVert  \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)} \rVert_{\calH_K}}_{=:\text{discrepancy}\bigl(\{\bsx\}_{i=0}^{n-1},K \bigr)}
	\, \underbrace{\inf_{c \in \RR}\lVert  f - c \rVert_{\calH_K}}_{=:\text{variation}(f,K)} \qquad \forall f \in \calH_K,  \label{eq:cuberrbd}
	\intertext{since $\hat{\mu}_n$ is exact for constants.  (It is assumed that $\calH$ contains constant functions so that $f \in \calH_K$ implies that $f - c \in \calH_K$.)  The squared norm of the error funtional can be expressed explicitly via \eqref{eq:sqnormL} as}
	\nonumber
	\text{discrepancy}^2\bigl(\{\bsx\}_{i=0}^{n-1},K \bigr) & = \lVert  \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)} \rVert_{\calH_K}^2 \\
	\nonumber
	& =
	\bigl(\mu(\cdot\cdot) - \hat{\mu}_n(\cdot\cdot)\bigr) \Bigl(\bigl(\mu(\cdot) - \hat{\mu}_n(\cdot)\bigr)\bigl( K(\cdot,\cdot \cdot) \bigr) \Bigr) \\
	\nonumber
	& = \int_{[0,1)^d \times [0,1)^d} K(\bst,\bsx) \, \mathrm{d} \bst \, \mathrm{d} \bsx  -\frac 2n  \sum_{i=0}^{n-1} \int_{[0,1)^d} K(\bst,\bsx_i) \, \mathrm{d} \bst \\
	& \qquad \qquad + \frac{1}{n^2} \sum_{i,j=0}^{n-1}  K(\bsx_i,\bsx_j). \label{eq:sqdisc}
\end{align}

Error bound \eqref{eq:cuberrbd} has two factors:
\begin{itemize}
	\item the discrepancy, which is the norm of the error functional, is a measure of deficiency of the node set, $\{\bsx\}_{i=0}^{n-1}$, and
	\item the variation is a measure of the roughness of the function defining our random variable whose expectation we wish to compute.
\end{itemize}
In general, designers of qMC methods want to construct sets or sequences of nodes with  discrepancy as small as possible, either by increasing $n$, if the computational budget allows, or by better placement of the nodes.  The constructions described in Section \ref{sec:construct} are ways of ensuring better placement of the nodes, provided that the generators are chosen well.  Practitioners of qMC want to formulate $\mu$ in a way to make the variation as small as possible.  This discussed in Section \ref{sec:reformulate}.

Although error bound \eqref{eq:cuberrbd} is elegant, it leaves several matters unresolved:
\begin{itemize}
	\item The reproducing kernel $K$ defines both the discrepancy and the variation.  This means that differnt choices of $K$ will lead to different error bounds for the same $\hat{\mu}_n$, even though the error is unchanged.
	\item The Hilbert space $\calH_K$ contains ones assumptions about $f$, such as smoothness or periodicity. Knowing $K$, however, does not automatically lead to an explict formula for the inner product of the associated Hilbert space, $\langle \cdot, \cdot \rangle_{\calH_K}$.  Conversely, having an explict formula for  $\langle \cdot, \cdot \rangle_{\calH_K}$ does not automatically lead to an explict formula for the reproducing kernel, $K$.  In both cases educated guesswork is involved.
	\item Choosing a more restrictive $\calH_K$ may lead to a sharper error bound, however, it is often infeasible in practice to check whether the ``black box'' $f$ lies in $\calH_K$.
	\item   Moreover, even if one is confident that $f \in \calH_K$, it is usually impractical to compute variation$(f,K)$, as we shall see below.  Error bound \eqref{eq:cuberrbd} cannot be used as a stopping criterion to determine the $n$ needed to satisfy an error tolerance.
\end{itemize}
In spite of these drawbacks, \eqref{eq:cuberrbd} is quite useful:
\begin{itemize}
	\item There are a family of commonly used reproducing kernels \cite{BerT-A04,Hic97a,Hic98b,Hic99b}. Once $K$ has been fixed, the discrepancy may be used as a comparison measure among different LD sequences, and its rate of decay---which sometimes can be computed theoretically---indicates how quickly $\hat{\mu}_n$ approaches $\mu$.
	\item When an explict formula for the inner product of the associated Hilbert space, $\langle \cdot, \cdot \rangle_{\calH_K}$ can be deduced, it suggests how to $\mu$ to reduce variation$(f)$.
\end{itemize}

@@

\section{Stopping Criteria} \label{sec:stop}
Art

@@

\bibliographystyle{spmpsci}
\bibliography{FJH23,FJHown23}

>>>>

When optimizing LD nodes, one requires an objective function which in some sense must assesses the quality of the LD nodes themselves, or the expected performance when implemented in a qMC estimate.  We will refer to such an objective function as a figure-of-merit (FOM) and popular choices include discrepancy or worst-case error metrics (see Section \ref{sec:discrepancy}) and spectral test... [come back to this]

\vspace{2mm}
\noindent
\textbf{Combinatorial Optimization.}
For several decades, combinatorial optimization techniques have been employed to obtain good generating parameters for LD constructions. For example, in the case of a lattice rule which is fully determined by the generating vector \(\boldsymbol{h} \in \mathbb{Z}_n^d \coloneqq \{0, 1, \ldots, n-1\}^d\) given $n$ and $d$, 
%While good generating vectors are known for \(d \leq 2\) \cite{DicEtal22a,SloJoe94}, the same does not hold for \(d \geq 3\), 
one often resorts to finding good $\boldsymbol{h}$ via computer search to minimize a chosen figure-of-merit. An exhaustive search is infeasible due to the exponential growth of the search space with \(d\). Thus, the component-by-component (CBC) construction, introduced by Korobov \cite{kor63} and later revisited by Sloan and collaborators \cite{Slo02}, reduces the search space to size \(dn\) by constructing \(\boldsymbol{h}\) one component at a time while keeping the previous components fixed. This greedy procedure also applies to finding good generating matrices $\mathsf{C}_j$ for digital sequences.
%Further simplification is possible by restricting the search to Korobov lattices of the form \(\boldsymbol{h} = (1, h, h^2, \ldots, h^{d-1}) \in \mathbb{Z}^d\). 
For certain choices of function spaces, there exist fast component-by-component construction implementations using fast Fourier transforms (FFTs) as the primary tool. We refer to \cite{DicKuo04a,JoeKuo03,KuoJoe02b} and references therein for works on the CBC construction for lattice rules and digital sequences, \cite{NuyCoo06a,NuyCoo06b} for their fast implementations and \cite{LatNet} for a QMC software tool called LatNet Builder by L'Ecuyer to generate LD nodes which relies heavily upon the CBC construction method.

\vspace{2mm}
\noindent
\textbf{Global Optimization.} Global optimization of LD nodes is often challenging due to the non-convex nature of the FOM as a function of the nodes. Traditional global optimization methods, such as regular gradient descent, struggle in this setting becoming trapped in suboptimal local minima and failing to identify the global optimum. One of the first works to globally optimize LD nodes, albeit a heuristic approach, was by Winker and Fang \cite{WinFan97b} employing the threshold accepting optimization algorithm. 
More recently, Rusch, Kirk, Bronstein, Lemieux and Rus \cite{ruschkirk24} developed the first machine-learning assisted optimization tool for LD node optimization in the Message-Passing Monte Carlo (MPMC) method. MPMC implements tools for geometric deep learning imposing a computational graph onto an initialized IID node set and learning a transformation mapping

\vspace{2mm}
\noindent
\textbf{Other Notable Methods.} Recently, a method called subset selection is formulated to choose from an $n−$element point set, the $k<n$ points which yield the smallest discrepancy. An exact selection algorithm was presented in \cite{cle22}, while a swap-based heuristic approach was used in \cite{cle24}. Furthermore, a method to generate optimal star-discrepancy point sets for fixed N and d based on a non-linear programming approach was suggested in \cite{}.

@@

\subsection{Digital Scrambles} \label{sec:scrambles}
For digital nets one can randomize even further by a scrambling that preserves the $t$-value.  Owen \cite{Owe95} proposed a scrambling of nets called nested uniform scrambling.  A simpler version called linear matrix scrambling was proposed by Matou\v{s}ek \cite{Mat98} and is described here.  Starting from the formulation of digital sequences involving generator matrices in \ref{eq:digitalB}, we multiply each generator matrix on the left by a lower triangular matrix with ones along the diagonals and elements below the diagonal that are randomly chosen to be $0$ or $1$ with equal probability:
\begin{subequations} \label{eq:scr_lattice}
\begin{gather} \label{eq:digitalscr}
	\begin{pmatrix} x^{\text{scr}}_{ij1} \\ x^{\text{scr}}_{ij2} \\ \vdots \end{pmatrix}
	= \mathsf{L_j} \mathsf{C}_j \begin{pmatrix} i_0 \\ i_1 \\ \vdots \end{pmatrix} + \begin{pmatrix} \Delta_{1j} \\ \Delta_{2j} \\ \vdots \end{pmatrix} \quad \pmod{2}, \qquad j = 1,\ldots, d, \ i = 0, 1, \ldots, \\
	\mathsf{L_j} :=
	\begin{pmatrix}
		1 & 0 & 0 & 0 & \cdots \\
		l_{21} & 1 & 0 & 0 & \cdots \\
		l_{31} & l_{32} & 1 & 0 & \cdots \\
		\vdots & \vdots & \vdots & \ddots & \ddots
	\end{pmatrix}, \qquad l_{\ell j} \overset {\text{IID}}{\sim} \calU\{0,1\}
\end{gather}
\end{subequations}
A random digital shift, $\bsDelta$ is also added, where $\Delta_{\ell j}$ denotes the $\ell^{\text{th}}$ digit of the $j^{\text{th}}$ dimension of the shift.

@@

As Halton points can be viewed as digital nets with different bases for each dimension, the randomization techniques used for digital nets can be applied to Halton pointsets. Random digital shifts for Halton set $x_{ij\ell} \mapsto (x_{ij\ell} + \Delta_{j\ell}) \mod b_j$ where the shift $\Delta_{j\ell} \in \{0,\dots,b_j-1\}$. More generally, a permutation scramble for digital nets sets $x_{ij\ell} \mapsto \Pi_{j\ell}(x_{ij\ell})$ where $\Pi_{j\ell}$ are independent random permutations of $\{0,\dots,b_j-1\}$. Random nested uniform scrambling and linear matrix scrambling have also been considered for Halton pointsets \cite{owen2024gain}.

@@

Instead of explicitly searching for good generating vectors or matrices, one can instead sample a few random generating vectors or generating matrices and then prune out bad choices \cite{}.  Random generating vectors $\bsh$ take $h_j \sim \calU\{0,\dots,n-1\}$. Random generating matrices for base $2$ digital nets set $c_{ij\ell} \sim \calU\{0,1\}$ for $i<\ell$ while enforcing $c_{iji}=1$ and $c_{ij\ell}=0$ for $i>\ell$. One idea is to generate

\section{Discrepancy} \label{sec:discrepancy}
So far, we have relied on eye tests and the Keister example to show how LD sequences are better than IID.  This section introduces the theory that shows why minimizing discrepancy leads to better approximations to $\mu$.

@@

Having a reproducing kernel Hilbert space with a known kernel $K$ allows us to derive a rigorous error bound for $\mu - \hat{\mu}_n$.  First note that for any bounded, linear functional, $L$ on the Hilbert space $\calH_K$, there is some $\eta_L \in \calH_K$ such that
\begin{equation*}
L(f) = \langle \eta_L , f \rangle_{\calH_K} \qquad \forall f \in \calH_K.
\end{equation*}
This is guaranteed by the Riesz Representation Theorem, and $\eta_L$ is called the representer of $L$.  Using the reproducing property of $K$, one may derive an explicit formula for $\eta_L(\bsx)$, namely,
\begin{equation*}
\eta_L(\bsx) = \langle K(\cdot, \bsx), \eta_L \rangle_{\calH_K} = L\bigl( K(\cdot,\bsx) \bigr).
\end{equation*}
From this expression for $\eta_L(\bsx)$ one may then calculate the squared norm of the linear functional $L$ as the squared norm of its representer:
\begin{equation} \label{eq:sqnormL}
	\lVert \eta_L \rVert_{\calH_K}^2 = \langle\eta_L, \eta_L \rangle_{\calH_K} = L(\eta_L) = L^{\cdot\cdot} \Bigl(L^{\cdot}\bigl( K(\cdot,\cdot \cdot) \bigr) \Bigr).
\end{equation}

If $\mu(\cdot)$ is a bounded linear functional on $\calH_K$, we may use the argument above to derive our error bound.  First we write the dependence of $\mu - \hat{\mu}_n$ on  $f$ explicitly as $\mu(f) - \hat{\mu}_n(f)$ and note that the error functional, $\mu(\cdot) - \hat{\mu}_n(\cdot)$, is  linear.  The Riesz Representation Theorem and the Cauchy-Schwarz inequality imply a tight error bound:
\begin{align}
	\nonumber
	\lvert\mu(f) - \hat{\mu}_n(f) \rvert
	& = \lvert \langle \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)}, f \rangle_{\calH_K} \rvert
	 \le \lVert  \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)} \rVert_{\calH_K} \, \lVert f \rVert_{\calH_K}\\
	\lvert\mu(f) - \hat{\mu}_n(f) \rvert
	& \le \underbrace{\lVert  \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)} \rVert_{\calH_K}}_{=:\text{discrepancy}\bigl(\{\bsx\}_{i=0}^{n-1},K \bigr)}
	\, \underbrace{\inf_{c \in \RR}\lVert  f - c \rVert_{\calH_K}}_{=:\text{variation}(f,K)} \qquad \forall f \in \calH_K,  \label{eq:cuberrbd}
	\intertext{since $\hat{\mu}_n$ is exact for constants.  (It is assumed that $\calH$ contains constant functions so that $f \in \calH_K$ implies that $f - c \in \calH_K$.)  The squared norm of the error funtional can be expressed explicitly via \eqref{eq:sqnormL} as}
	\nonumber
	\text{discrepancy}^2\bigl(\{\bsx\}_{i=0}^{n-1},K \bigr) & = \lVert  \eta_{\mu(\cdot) - \hat{\mu}_n(\cdot)} \rVert_{\calH_K}^2 \\
	\nonumber
	& =
	\bigl(\mu(\cdot\cdot) - \hat{\mu}_n(\cdot\cdot)\bigr) \Bigl(\bigl(\mu(\cdot) - \hat{\mu}_n(\cdot)\bigr)\bigl( K(\cdot,\cdot \cdot) \bigr) \Bigr) \\
	\nonumber
	& = \int_{[0,1)^d \times [0,1)^d} K(\bst,\bsx) \, \mathrm{d} \bst \, \mathrm{d} \bsx  -\frac 2n  \sum_{i=0}^{n-1} \int_{[0,1)^d} K(\bst,\bsx_i) \, \mathrm{d} \bst \\
	& \qquad \qquad + \frac{1}{n^2} \sum_{i,j=0}^{n-1}  K(\bsx_i,\bsx_j). \label{eq:sqdisc}
\end{align}

Error bound \eqref{eq:cuberrbd} has two factors:
\begin{itemize}
	\item the discrepancy, which is the norm of the error functional, is a measure of deficiency of the node set, $\{\bsx\}_{i=0}^{n-1}$, and
	\item the variation is a measure of the roughness of the function defining our random variable whose expectation we wish to compute.
\end{itemize}
In general, designers of qMC methods want to construct sets or sequences of nodes with  discrepancy as small as possible, either by increasing $n$, if the computational budget allows, or by better placement of the nodes.  The constructions described in Section \ref{sec:construct} are ways of ensuring better placement of the nodes, provided that the generators are chosen well.  Practitioners of qMC want to formulate $\mu$ in a way to make the variation as small as possible.  This is discussed in Section \ref{sec:reformulate}.

Although error bound \eqref{eq:cuberrbd} is elegant, it leaves several matters unresolved:
\begin{itemize}
	\item The reproducing kernel $K$ defines both the discrepancy and the variation.  This means that differnt choices of $K$ will lead to different error bounds for the same $\hat{\mu}_n$, even though the error is unchanged.
	\item The Hilbert space $\calH_K$ contains ones assumptions about $f$, such as smoothness or periodicity. Knowing $K$, however, does not automatically lead to an explicit formula for the inner product of the associated Hilbert space, $\langle \cdot, \cdot \rangle_{\calH_K}$.  Conversely, having an explicit formula for  $\langle \cdot, \cdot \rangle_{\calH_K}$ does not automatically lead to an explicit formula for the reproducing kernel, $K$.  In both cases educated guesswork is involved.
	\item Choosing a more restrictive $\calH_K$ may lead to a sharper error bound, however, it is often infeasible in practice to check whether the ``black box'' $f$ lies in $\calH_K$.
	\item   Moreover, even if one is confident that $f \in \calH_K$, it is usually impractical to compute variation$(f,K)$, as we shall see below.  Error bound \eqref{eq:cuberrbd} cannot be used as a stopping criterion to determine the $n$ needed to satisfy an error tolerance.
\end{itemize}
In spite of these drawbacks, \eqref{eq:cuberrbd} is quite useful:
\begin{itemize}
	\item There are a family of commonly used reproducing kernels \cite{BerT-A04,Hic97a,Hic98b,Hic99b}. Once $K$ has been fixed, the discrepancy may be used as a comparison measure among different LD sequences, and its rate of decay---which sometimes can be computed theoretically---indicates how quickly $\hat{\mu}_n$ approaches $\mu$.
	\item When an explicit formula for the inner product of the associated Hilbert space, $\langle \cdot, \cdot \rangle_{\calH_K}$, can be deduced, it suggests how to formulate $\mu$ to reduce variation$(f)$.
\end{itemize}

@@

\section{Stopping Criteria} \label{sec:stop}

Art

@@

\bibliographystyle{spmpsci}
\bibliography{FJH23,FJHown23,main}
